% Copyright (C) 2014-2020 by Thomas Auzinger <thomas@auzinger.name>
% Add oneside to make a more read-friendly pdf
\documentclass[draft, final, x11names, svgnames]{template/vutinfth} % Remove option 'final' to obtain debug information.

\usepackage[sorting=nyt, style=alphabetic, backend=biber, alldates=iso]{biblatex}
\usepackage{amsmath}    % Extended typesetting of mathematical expression.
\usepackage{amssymb}    % Provides a multitude of mathematical symbols.
%\usepackage{mathtools}  % Further extensions of mathematical typesetting.
\usepackage{fontspec}
\usepackage{microtype}  % Small-scale typographic enhancements.
\usepackage[inline]{enumitem} % User control over the layout of lists (itemize, enumerate, description).
\usepackage{multirow}   % Allows table elements to span several rows.
\usepackage{booktabs}   % Improves the typesettings of tables.
\usepackage{subcaption} % Allows the use of subfigures and enables their referencing.
\usepackage[ruled,linesnumbered,algochapter]{algorithm2e} % Enables the writing of pseudo code.
\usepackage[usenames,dvipsnames,table]{xcolor} % Allows the definition and use of colors. This package has to be included before tikz.
\usepackage[autostyle,german=guillemets]{csquotes}
\usepackage{minted}
\usepackage{nag}       % Issues warnings when best practices in writing LaTeX documents are violated.
\usepackage{todonotes} % Provides tooltip-like todo notes.
\usepackage{titlesec}
\usepackage{multirow}
% For line breaks in \texttt
% \usepackage[htt]{hyphenat}
% Maybe remove this line for final digital export
\usepackage[hidelinks]{hyperref}
\usepackage{hyperref}  % Enables cross linking in the electronic document version. This package has to be included second to last.
\usepackage[acronym,toc]{glossaries} % Enables the generation of glossaries and lists fo acronyms. This package has to be included last.

\addbibresource{references.bib}

\setmainfont{LinBiolinum_R.otf}[
    Path = /usr/local/texlive/2020/texmf-dist/fonts/opentype/public/libertine/,
    BoldFont = LinBiolinum_RB.otf,
    ItalicFont = LinBiolinum_RI.otf
]

\setmonofont[Scale=0.8]{JetBrainsMono-Variable.ttf}[
    Path = /usr/local/share/fonts/
]

% Hyphenation
\hyphenation{Web-As-sem-bly}

\definecolor{friendlybg}{HTML}{f0f0f0}
\newcommand{\inl}[1]{\colorbox{friendlybg}{\texttt{#1}}}

% Quotes
\newcommand{\quot}[1]{\foreignquote{german}{#1}}

\definecolor{BrightPurple}{HTML}{8c14ee}
\titleformat{\section}
    {\Large\normalfont\bfseries}  % Large, bold, sans serif font for section
    {}                          % No format applied to whole title
    {0pt}                       % No separation between label and title
    {\thesection~·~}            % Start with section number
    [{\color{BrightPurple}\hrule}]      % Underline with colored ruler

% Define convenience functions to use the author name and the thesis title in the PDF document properties.
\newcommand{\authorname}{Philipp Gackstatter} % The author name without titles.
\newcommand{\thesistitle}{A WebAssembly Container Runtime for Serverless Edge Computing} % The title of the thesis. The English version should be used, if it exists.

% Set PDF document properties
\hypersetup{
    pdfpagelayout   = TwoPageRight,           % How the document is shown in PDF viewers (optional).
    linkbordercolor = {Melon},
    pdfauthor       = {\authorname},
    pdftitle        = {\thesistitle},
    pdfsubject      = {Serverless WebAssembly Container Runtime},
    pdfkeywords     = {Serverless,
                       Function-as-a-Service,
                       WebAssembly,
                       Container Runtime,
                       Edge Computing}
}

% Switch to arabic numbering for the entire document
% \renewcommand{\pagenumbering}[1]{}

\setpnumwidth{2.5em}        % Avoid overfull hboxes in the table of contents (see memoir manual).
\setsecnumdepth{subsection} % Enumerate subsections.

\nonzeroparskip             % Create space between paragraphs (optional).
\setlength{\parindent}{0pt} % Remove paragraph identation (optional).

\makeindex      % Use an optional index.
\makeglossaries % Use an optional glossary.
%\glstocfalse   % Remove the glossaries from the table of contents.

\setauthor{}{\authorname}{BSc.}{male}
\setadvisor{Univ.Prof. Mag.rer.soc.oec. Dr.rer.soc.oec.}{Schahram Dustdar}{}{male}

% For bachelor and master theses:
\setfirstassistant{Univ.Ass.}{Pantelis Frangoudis}{PhD}{male}
%\setsecondassistant{Pretitle}{Forename Surname}{Posttitle}{male}
%\setthirdassistant{Pretitle}{Forename Surname}{Posttitle}{male}


\setregnumber{11846040}
\setdate{\day}{\month}{\year} % Set date with 3 arguments: {day}{month}{year}.
\settitle{\thesistitle}{Eine WebAssembly Container Runtime für Serverless Edge Computing}
% Sets English and German version of the title (both can be English or German). If your title contains commas, enclose it with additional curvy brackets (i.e., {{your title}}) or define it as a macro as done with \thesistitle.

%\setsubtitle{Optional Subtitle of the Thesis}{Optionaler Untertitel der Arbeit} % Sets English and German version of the subtitle (both can be English or German).

\setthesis{master}
\setmasterdegree{dipl.}
\setcurriculum{Software Engineering \& Internet Computing}{Software Engineering \& Internet Computing}

\begin{document}

\frontmatter % Switches to roman numbering.
% The structure of the thesis has to conform to the guidelines at
%  https://informatics.tuwien.ac.at/study-services

\addtitlepage{naustrian} % German title page (not for dissertations at the PhD School).
\addtitlepage{english} % English title page.
\addstatementpage

\begin{danksagung*}
\todo{Ihr Text hier.}
\end{danksagung*}

\begin{acknowledgements*}
\todo{Enter your text here.}
\end{acknowledgements*}

\begin{kurzfassung}

Serverless Computing ist durch die Abstraktion des Infrastrukturmanagements ein beliebter Teil des Cloud Computing geworden. Es ermöglicht automatisch skalierende Funktionen zu schreiben, wobei nur die verbrauchte Rechenzeit in Rechnung gestellt wird. Dieses Modell ist ideal um schwallartige Anfragen zu verarbeiten, die etwa aus dem Internet der Dinge kommen. Allerdings können selbst die räumlich nahen Edge Computer latenzsensible Anwendungsfälle aufgrund eines signifikanten Performanceproblems nicht bedienen.

Beim ersten Aufruf einer Funktion muss die Ausführungsumgebung gestartet werden -- üblicherweise ein Docker Container. Dieser \emph{Kaltstart} benötigt mehrere hundert Millisekunden bis hin zu mehreren Sekunden bei parallelen Anfragen, was Serverless in Kombination mit Docker für diese Anwendungsfälle ungeeignet macht. Der Betrieb dieser Plattformen auf Edge Computern mit begrenzten Ressourcen erhöht die Kaltstartzeit noch weiter.

Um dieses Problem zu beheben verfolgen wir einen radikalen Ansatz, bei dem Docker, die de-facto Standard-Runtime für Serverless Computing, gänzlich durch eine leichtgewichtigere Technologie ersetzt wird. Jede Alternative muss eine ähnliche Sicherheit, Sprachunabhängigkeit und Performance bieten. WebAssembly ist eine aufkommende Technologie die diese Eigenschaften für sich beansprucht. In dieser Arbeit untersuchen wir die Eignung von WebAssembly für den Einsatz in einer Serverless Container Runtime für Apache OpenWhisk, einem Serverless Framework, mit einem Fokus auf Edge Computing. Wir wählen drei WebAssembly Runtimes um unsere Container Runtime darauf aufzubauen und stellen dessen Design vor, um schnelle Kaltstarts sowie eine hohe Performance zu erzielen.

Wir vergleichen OpenWhisk in der Dockervariante mit den drei WebAssembly Container Runtimes im Hinblick auf Kaltstartzeit, Performance und Speicherverbrauch in umfangreichen Experimenten. Auf einem Raspberry Pi, einem typischen low-end Edge Computer, erreichen unsere WebAssembly Container Runtimes einen 99.5\% schnelleren Kaltstart als Docker und einen 94\% schnelleren Start auf Serverhardware. Docker's Performance wird stark von der Kaltstartzeit beeinträchtigt, während der Durchsatz von WebAssembly Containern nicht spürbar davon beeinflusst wird. In einem Lasttest mit CPU- und I/O-gebundenen Funktionen erreichen unsere Container Runtimes mindestens 2,4 und bis zu 4,2 mal mehr Durchsatz als Docker. Insgesamt zeigt sich WebAssembly als geeignete Alternative für Serverless Plattformen durch den leichtgewichtigen Isolationsmechanismus, der schnelle Starts ermöglicht, sowie Unterstützung für vorzeitige Kompilierung zu schnellem nativen Maschinencode. Unser Ansatz ist daher ein aussichtsreicher Schritt hin zu latenzarmem Serverless Edge Computing.

\end{kurzfassung}

\begin{abstract}

Serverless computing has become a popular part of the cloud computing model, thanks to abstracting away infrastructure management. It enables any developer to write functions that auto-scale, while only paying for the used compute time. This model is ideal for handling the unpredictable and bursty workloads in the Internet of Things. However, even the physically close edge computers cannot facilitate low latency use cases due to a significant performance issue.

The first invocation of a function requires creating its execution environment -- typically a Docker container. This \emph{cold start} can take hundreds of milliseconds or even multiple seconds under concurrent requests, making serverless with Docker unsuitable for these use cases. Running these platforms on edge devices with limited CPU power and memory available, increases cold start latencies even further.

To alleviate this issue, we follow a radical approach and replace Docker, the de-facto standard runtime for serverless computing, with a lighter-weight technology. However, any alternative must offer similar levels of security, language-agnosticism and high performance. A nascent technology, which claims to have these properties, is a target format called WebAssembly. In this work, we examine WebAssembly's suitability for use in a serverless container runtime for Apache OpenWhisk, a serverless framework, with a focus on edge computing.
We choose three WebAssembly runtimes to build our container runtime on top of and present its design to achieve fast cold starts while retaining a high performance.

We compare the Docker-based OpenWhisk with the three WebAssembly container runtimes in terms of cold start time, execution performance and memory usage via extensive experiments in a testbed. On a Raspberry Pi, a typical low-end edge computing device, we find our WebAssembly container runtimes to achieve 99.5\% faster cold starts than Docker, and 94\% faster starts on server-grade hardware. Docker's execution performance is strongly linked to the cold start time, while the throughput of WebAssembly containers is not noticeably susceptible to cold starts. In a mixed CPU- and I/O-bound workload, our container runtimes achieve at least 2.4 and up to 4.2 times the throughput of Docker. Overall, we find WebAssembly to be very suitable for use in serverless platforms due to its lightweight isolation mechanism, enabling fast startups, as well as support for ahead-of-time compilation to fast, native code. Our approach is thus a promising step towards low-latency serverless edge computing.

\end{abstract}

% Select the language of the thesis, e.g., english or naustrian.
\selectlanguage{english}

% Sets the depth of the toc
\setcounter{tocdepth}{2}
% Add a table of contents (toc).
\tableofcontents % Starred version, i.e., \tableofcontents*, removes the self-entry.

% Switch to arabic numbering and start the enumeration of chapters in the table of content.
\mainmatter

\input{chapters/introduction.tex}
\input{chapters/background.tex}
\input{chapters/design.tex}
\input{chapters/evaluation.tex}
\input{chapters/relatedwork.tex}
\input{chapters/conclusion.tex}

\backmatter

% Use an optional list of figures.
\listoffigures % Starred version, i.e., \listoffigures*, removes the toc entry.

% Use an optional list of tables.
\cleardoublepage % Start list of tables on the next empty right hand page.
\listoftables % Starred version, i.e., \listoftables*, removes the toc entry.

\cleardoublepage
\listoflistings

% Add an index.
\printindex

% Add a glossary.
\printglossaries

% Add a bibliography.
\begin{sloppypar}
    \printbibliography[title={References}, nottype=online]
    \printbibliography[title={Online Resources}, type=online]
\end{sloppypar}

\end{document}
